# Two approaches, one object

There are two main ways to initiate a plotly object in R. The `plot_ly()` function transforms _data_ into a plotly object, while the `ggplotly()` function transforms a _ggplot object_ into a plotly object [@ggplot2]; [@plotly]. Regardless how plotly object is created, printing it results in an interactive web-based visualization with tooltips, zooming, and panning enabled by default. It is also possible to enable more advanced interactive techniques, such as [animation](#animation) and [linked highlighting](#linked-highlighting).

The initial inspiration for the `plot_ly()` function was to support [plotly.js](https://github.com/plotly/plotly.js) chart types that **ggplot2** doesn't support, such as 3D surface and mesh plots. Over time, this effort snowballed into a full-blown interface to the entire plotly.js graphing library with additional abstractions inspired by the grammar of graphics [@Wilkinson:2005]. This "non-ggplot2" interface to plotly.js is (somewhat intentionally) not as fully featured as **ggplot2** to provide more efficient alternative for certain plots. Also, by having high-level functions that can modify plotly objects, we're also able to [extend `ggplotly()`](#extend-ggplotly). In the next section, we introduce some similarities and differences between `ggplotly()` and `plot_ly()` through a case study.

## A case study of housing sales in Texas

The **plotly** package depends on **ggplot2** which bundles a data set on monthly housing sales in Texan cities acquired from the [TAMU real estate center](http://recenter.tamu.edu/). After the loading the package, the data is "lazily loaded" into your session, so you may reference it by name:

```{r}
library(plotly)
txhousing
```

In attempt to understand house price behavior over time, we could plot `date` on x, `median` on y, and group the lines connecting these x/y pairs by `city`. Using **ggplot2**, we can _initiate_ a ggplot object with the `ggplot()` function which accepts a data frame and a mapping from data variables to visual aesthetics. By just initiating the object, **ggplot2** won't know how to geometrically represent the mapping until we add a layer to the plot via one of `geom_*()` (or `stat_*()`) functions (in this case, we want `geom_line()`). In this case, it is also a good idea to specify alpha transparency so that 5 lines plotted on top of each other appear as solid black, to help avoid overplotting.

```{block, type='rmdtip'}
If you're new to **ggplot2**, the [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf) provides a nice quick overview. The [online docs](http://docs.ggplot2.org/current/) or [R graphics cookbook](http://www.cookbook-r.com/Graphs/) are helpful for learning by example, and the [ggplot2 book](https://github.com/hadley/ggplot2-book) provides a nice overview of the conceptual underpinnings.
```

```{r}
p <- ggplot(txhousing, aes(date, median)) +
  geom_line(aes(group = city), alpha = 0.2)
```

Now that we have a valid **ggplot2** object, `p`, the **plotly** package provides the `ggplotly()` function which converts a ggplot object to a plotly object. By default, it supplies the entire aesthetic mapping to the tooltip, but the `tooltip` argument provides a way to restrict tooltip info to a subset of that mapping. Furthermore, in cases where the statistic of a layer is something other than the identity function (e.g., `geom_bin2d()` and `geom_hex()`), relevant "intermediate" variables generated in the process are also supplied to the tooltip. This provides a nice mechanism for decoding visual aesthetics (e.g., color) used to represent a measure of interest (e.g, count/value). In Figure \@ref(fig:ggsubplot), the `subplot()` function from the **plotly** package (discussed in more detail in [subplots](#subplot)), which accepts a collection of ggplot and/or plotly objects, helps to concisely display tooltips for a number of scenarios, and how to suppress them.

```{r ggsubplot, fig.cap = "By default all aesthetic mappings are displayed upon hover. The tooltip argument provides a way to display a subset of the mapping."}
subplot(
  p, ggplotly(p, tooltip = "city"), 
  ggplot(txhousing, aes(date, median)) + geom_bin2d(),
  ggplot(txhousing, aes(date, median)) + geom_hex(),
  nrows = 2, shareX = TRUE, shareY = TRUE,
  titleY = FALSE, titleX = FALSE
)
```

```{block, type='rmdtip'}
Although **ggplot2** does not have a `text` aesthetic, the `ggplotly()` function recognizes this aesthetic and displays it in the tooltip by default. In addition to providing a way to supply "meta" information, it also provides a way to customize your tooltips (do this by restricting the tooltip to the text aesthetic -- `ggplotly(p, tooltip = "text")`)
```

The `ggplotly()` function translates most things that you can do in **ggplot2**, but not quite everything. To help demonstrate the coverage, I've built a [plotly version of the ggplot2 docs](http://ropensci.github.io/plotly/ggplot2). This version of the docs displays the `ggplotly()` version of each plot in a static form (to reduce page loading time), but you can click any plot to view its interactive version. The next section deomnstrates how to create plotly.js visualizations via the R package, without **ggplot2**, via [the `plot_ly()` interface](#plot_ly). We'll then leverage those concepts to [extend `ggplotly()`](#extend-ggplotly).

### The `plot_ly()` interface {#plot_ly}

The cognitive framework underlying the `plot_ly()` interface draw inspiration from the layered grammar of graphics [@Wickham:2010hya], but in contrast to `ggplotly()`, it provides a more flexible and direct interface to [plotly.js](https://github.com/plotly/plotly.js). It is more direct in the sense that it doesn't call **ggplot2**'s sometimes expensive plot building routines, and it is more flexible in the sense that data frames are not required, which is useful for visualizing matrices, as shown in [Get Started](#get-started). Although data frames are not required, it is recommended to use them whenever possible, especially when constructing a plot with multiple layers or groups. 

When a data frame is associated with a **plotly** object, it allows us to manipulate the data underlying that object in the same way we would directly manipulate the data. Currently, `plot_ly()` borrows semantics from and provides special plotly methods for generic functions in the **dplyr** and **tidyr** packages [@dplyr]; [@tidyr]. Most importantly, `plot_ly()` recognizes and preserves groupings created with **dplyr**'s `group_by()`.

```{r}
library(dplyr)
tx <- group_by(txhousing, city)
# initiate a plotly object with date on x and median on y
p <- plot_ly(tx, x = ~date, y = ~median)
# plotly_data() returns data associated with a plotly object, note the group attribute!
plotly_data(p)
```

Defining groups in this fashion ensures `plot_ly()` will produce at least one graphical mark per group.^[In practice, it's easy to forget about "lingering" groups (e.g., `mtcars %>% group_by(vs, am) %>% summarise(s = sum(mpg))`), so in some cases, you may need to `ungroup()` your data before plotting it.] So far we've specified `x` and `y` attributes in the plotly object `p`, but we have not yet specified the geometric relation between these x/y pairs. Similar to `geom_line()` in **ggplot2**, the `add_lines()` function connects (a group of) x/y pairs with lines in the order of their `x` values, which is useful when plotting time series as shown in Figure \@ref(fig:houston).



Similar to the `ggplot()` function in **ggplot2**, `plot_ly()` initializes a plot object, and any attributes defined here . 

```{r houston, fig.cap = "Median house sales in houston compared to other Texan cities"}
# we get one line per city since p knows city is a grouping variable
p2 <- add_lines(p, alpha = 0.2, name = "Texan Cities", hoverinfo = "none")
# add another line to highlight houston
add_lines(
  p2, name = "Houston", color = I("red"),
  data = filter(txhousing, city == "Houston")
)
```

The **plotly** package has a collection of `add_*()` functions, all of which inherit attributes defined in `plot_ly()`. These functions also inherit the data associated with the plotly object provided as input, unless otherwise specified in the `data` argument. I prefer to think about `add_*()` functions like a layer in **ggplot2**, which is slightly different from the concept of a plotly.js trace. In Figure \@ref(fig:houston), there is a 1-to-1 correspondence between layers and traces, but as you will find in the [The plotly cookbook](the-plotly-cookbook), an `add_*()` function can produce numerous traces. 

In the first layer of Figure \@ref(fig:houston), there is one line per city, but this entire layer is implemented with a single trace. We _could have_ produced one trace for each line, but this is expensive because, by default, each trace produces a legend entry and tries to display meaningful hover information. It is much more efficient to render this layer as a single trace with missing values to differentiate groups. In fact, this is exactly how the group aesthetic is translated in `ggplotly()`; otherwise, layers with many groups (e.g., `geom_map()`) would be slow to render.

Since every **plotly** function modifies a plotly object (or the data underlying that object), we can express complex multi-layer plots as a sequence (or, more specifically, a directed acyclic graph) of data manipulations and mappings to the visual space. Moreover, **plotly** functions are designed to take a plotly object as input, and return a modified plotly object, making it easy to chain together operations via the pipe operator (`%>%`) from the **magrittr** package [@magrittr]. Consequently, we can re-express Figure \@ref(fig:houston) in a much more readable and understandable fashion.

```{r, eval = FALSE}
txhousing %>%
  group_by(city) %>%
  plot_ly(x = ~date, y = ~median) %>%
  add_lines(alpha = 0.2, name = "Texan Cities", hoverinfo = "none") %>%
  filter(city == "Houston") %>%
  add_lines(color = I("red"), name = "Houston")
```

Not only does it make you

```{r}
txhousing %>%
  group_by(city) %>%
  plot_ly(x = ~date, y = ~median) %>%
  add_lines(alpha = 0.2, name = "Texan Cities", hoverinfo = "none") %>%
  group_by(date) %>% 
  summarise(
    q1 = quantile(median, 0.25, na.rm = TRUE),
    m = median(median, na.rm = TRUE),
    q3 = quantile(median, 0.75, na.rm = TRUE)
  ) %>%
  add_lines(y = ~m, color = I("red"), name = "median") %>%
  add_ribbons(ymin = ~q1, ymax = ~q3, color = I("red"), name = "IQR")
```


TODO:
(1) example with data passed explicity to data argument
(2) example of multi layer plot without data frames




This "data-plot-pipeline" is desirable for a number of reasons: (1) makes your code easier to read and understand (2) any generic function that modifies a data frame can be incorporated into the pipeline (TODO: write a section on writing S3 methods?) (3) as discussed in [Linked Highlighting](#linked-highlighting), it provides a natural mechanism for implementing the pipelines necessary for interactive graphics [@plumbing].

Note that, similar to `ggplot()`, `plot_ly()` only initiatizes the plot with "global" visual attributes



This plot doesn't tell us much (other than prices seems to be increasing overall) since it's hard to see what's happening in individual cities. We could highlight a particular series by subsetting the data 



## Extending `ggplotly()`

```{r}
p <- ggplot(txhousing, aes(x = date, y = median)) +
  geom_line(aes(group = city), alpha = 0.3) +
  geom_smooth()

ggplotly(p) %>%
  filter(city == "Houston") %>%
  add_lines(color = I("red"), name = "Houston")
```

* Expose the pre-statistics data. 



## Choosing an interface {sec: choosing}


```{r}
library(plotly)
d <- subset(diamonds, carat <= 2)
p <- ggplot(d, aes(log(carat), log(price))) + 
  geom_hex() + 
  geom_smooth(method = "lm", colour = "red", se = FALSE)
ggplotly(p)
```



That's great, but there are a few reasons why this approach is limited:

1. ggplot2's interface wasn't designed for interactive graphics. Directly extending the grammar to support more advanced types of interaction (e.g., linked brushing) is a risky endeavor.
2. ggplot2 requires data frame(s) and can be inefficient (especially for time series).
3. ggplot2 does not have a functional interface (making it awkward to combine with modern functional interfaces such as dplyr), and does not satisfy [referential transparency](https://en.wikipedia.org/wiki/Referential_transparency) (making it easier to program with -- for more details, see )
4. `ggplotly()` tries to replicate _exactly_ what you see in the corresponding static ggplot2 graph. To do so, it sends axis tick information to plotly as [tickvals](https://plot.ly/r/reference/#layout-xaxis-tickvals)/[ticktext](https://plot.ly/r/reference/#layout-xaxis-ticktext) properties, and consequently, axis ticks do not update on zoom events.
