# Two interfaces, one central object

There are two main ways to create a **plotly** object (i.e., visualization) in R. The `plot_ly()` function transforms _data_ into a **plotly** object, while the `ggplotly()` function transforms a **ggplot2** object into a **plotly** object [@ggplot2]; [@plotly]. No matter which interface you use, printing a **plotly** object results in an interactive web-based visualization with tooltips, zooming, and panning. Furthermore, once you have a **plotly** object, any **plotly** function can be used to modify that **plotly** object. As a result, we can effectively modify **ggplot2** plots, and leverage features exclusive to plotly (see section XXX for some compelling examples showing the power of this idea).

A good question to ask now would be: "Why the need for two interfaces"? The quick and easy answer is that **ggplot2** can be quite slow; but has an elegant interface with many useful abstractions (e.g., a consistent interface for altering scales), smart defaults, and a very extensive set of features. On the other hand, `plot_ly()` can be much faster, mostly because it does less magic. Furthermore, **ggplot2** is focused on 2D graphics, whereas plotly.js has awesome 3D scientific visualization capabilities -- the inspiration having `plot_ly()` in the first place. Even though **ggplot2** does a lot of things right (which `plot_ly()` strives to adopt), one could argue it has a number of design flaws (which `plot_ly()` strives to avoid).

As shown explicitly in the next section, the design of the `plot_ly()` interface is heavily inspired by **ggplot2**, and it's implementation of the Grammar of Graphics [@Wilkinson:2005]. In its layered approach to the grammar of graphics [@Wickham:2010hya], it defines a _layer_ as the combination of five elements: (1) data (2) aesthetic mapping (3) geometry (4) statistical summary (5) positional adjustment. The fact that these elements are easily understood in isolation, but can be combined in many ways to produce complex visualizations, is a huge reason for ggplot2's success (not to mention its support for a bunch of different useful geometries and statistics). 

One could argue, however, that embedding data transformations (e.g., statistical summaries and positional adjustments) inside the plot building step was a mistake since other packages can't leverage that functionality. We're already seeing [glimpses](https://channel9.msdn.com/Events/useR-international-R-User-conference/useR2016/Towards-a-grammar-of-interactive-graphics) of Hadley's vision for the future where statistical computations and data transformations are decoupled from the actual rendering stage [@ggvis]; [@ggstat]. One could also argue that **ggplot2**'s overriding of the `+` operator to combine components is a design flaw -- layers should instead be described as a sequence of pure functions which perform data transformation(s) and mapping(s) to visuals.

As always, the devil is in the detail, so I've provided a list at the end of this chapter that lay out more of the limitations behind each interface which may be helpful for [Choosing an interface](#sec:choosing). For now, to help point out some obvious similarities and differences, we'll start by creating the same example with both interfaces.

## A case study of housing sales in Texas

The plotly package depends on ggplot2 which bundles a data set on monthly housing sales in Texan cities acquired from the [TAMU real estate center](http://recenter.tamu.edu/).

```{r}
library(plotly)
txhousing
```

In attempt to understand house price behavior over time, we could plot `date` on x, `median` on y, and group the lines connecting these x/y pairs by `city`. 

### The `ggplotly()` interface

```{r}
txhousing %>%
  group_by(city) %>%
  plot_ly(x = ~date, y = ~median) %>%
  add_lines(alpha = 0.3)
```

```{r}
library(plotly)
p <- ggplot(txhousing, aes(date, median, group = city)) +
  geom_line(alpha = 0.3)
ggplotly(p)
```

This plot doesn't tell us much (other than prices seems to be increasing overall) since it's hard to see what's happening in individual cities. We could highlight a particular series by subsetting the data 

```{r}
```

To compare the monthly median house price across cities

If you use ggplot2, `ggplotly()` adds interactivity (specifically, tooltips, zoom, and pan) to your plots with minimal effort. 


### The `plot_ly()` interface

The `plot_ly()` function draws inspiration from ggplot2's implementation of the grammar of graphics, but provides a more flexible and direct interface to [plotly.js](https://github.com/plotly/plotly.js). The interface is also functional, and designed to work with dplyr's generic data manipulation functions, so visualizations can be described as a sequence of data manipulations and visual components via the pipe operator (`%>%`) from the magrittr package.

```{r}
txhousing %>%
  group_by(city) %>%
  plot_ly(x = ~date, y = ~median) %>%
  add_lines(color = I("black"), alpha = 0.3)
```


```{r}
txhousing %>%
  group_by(city) %>%
  plot_ly(x = ~date, y = ~median) %>%
  add_lines(alpha = 0.3, color = I("black"), name = "Texan Cities") %>%
  filter(city == "Houston") %>%
  add_lines(color = I("red"), name = "Houston")
```


### Adding layers

The `ggplotly()` function translates [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html) graphics to a plotly equivalent, for example:

```{r}
library(plotly)
p <- qplot(data = txhousing, x = date, y = median, group = city, alpha = I(0.3), geom = "line")

p <- p + geom_line(data = subset(txhousing, city == "Houston"), color = "red")


ggplotly(p) %>%
  filter(city == "Houston") %>%
  add_lines(color = I("red"), name = "Houston")
```

## Combining the two

```{r}
p <- ggplot(txhousing, aes(x = date, y = median)) +
  geom_line(aes(group = city), alpha = 0.3) +
  geom_smooth()

ggplotly(p) %>%
  filter(city == "Houston") %>%
  add_lines(color = I("red"), name = "Houston")
```

* Expose the pre-statistics data. 


## TODO: Section on controlling tooltips and creating custom tooltip via `text`


## Choosing an interface {sec: choosing}


```{r}
library(plotly)
d <- subset(diamonds, carat <= 2)
p <- ggplot(d, aes(log(carat), log(price))) + 
  geom_hex() + 
  geom_smooth(method = "lm", colour = "red", se = FALSE)
ggplotly(p)
```



That's great, but there are a few reasons why this approach is limited:

1. ggplot2's interface wasn't designed for interactive graphics. Directly extending the grammar to support more advanced types of interaction (e.g., linked brushing) is a risky endeavor.
2. ggplot2 requires data frame(s) and can be inefficient (especially for time series).
3. ggplot2 does not have a functional interface (making it awkward to combine with modern functional interfaces such as dplyr), and does not satisfy [referential transparency](https://en.wikipedia.org/wiki/Referential_transparency) (making it easier to program with -- for more details, see )
4. `ggplotly()` tries to replicate _exactly_ what you see in the corresponding static ggplot2 graph. To do so, it sends axis tick information to plotly as [tickvals](https://plot.ly/r/reference/#layout-xaxis-tickvals)/[ticktext](https://plot.ly/r/reference/#layout-xaxis-ticktext) properties, and consequently, axis ticks do not update on zoom events.
