# The plotly cookbook

This chapter demonstrates the capabilities of `plot_ly()` through a series of examples. The `plot_ly()` function does provide a direct interface to plotly.js, so anything in [the figure reference](https://plot.ly/r/reference/) can be specified via `plot_ly()`, but this chapter will focus more on the semantics unique to the R package that can't be found on the figure reference. Along the way, we will touch on some best practices in visualization.

## Scatter traces

A plotly visualization is composed of one (or more) trace(s), and every trace has a `type`. The default trace type, "scatter", can be used to draw a large amount of geometries, and actually powers many of the `add_*()` functions such as `add_markers()`, `add_lines()`, `add_paths()`, `add_segments()`, `add_ribbons()`, and `add_polygons()`. These functions make assumptions about the [mode](https://plot.ly/r/reference/#scatter-mode) of the scatter trace (among other things), but any of the attributes listed under the [scatter section of the figure reference](https://plot.ly/r/reference/#scatter) may be passed to these functions.

### Scatterplots

The scatterplot is useful for visualizing the correlation between two quantitative variables. If you supply a numeric vector for x and y in `plot_ly()`, it defaults to a scatterplot, but you can also be explicit about adding a layer of markers/points via the `add_markers()` function. A common problem with scatterplots is overplotting, meaning that there are multiple observations occupying the same (or similar) x/y locations. There are a few ways to combat overplotting including: alpha transparency, hollow symbols, and [2D density estimation](). Figure \@ref(fig:scatterplots) shows three versions of a basic scatterplot:

```{r scatterplots, fig.cap = "Three versions of a basic scatterplot"}
subplot(
  plot_ly(mpg, x = ~cty, y = ~hwy, name = "default"),
  plot_ly(mpg, x = ~cty, y = ~hwy) %>% add_markers(alpha = 0.2, name = "alpha"),
  plot_ly(mpg, x = ~cty, y = ~hwy) %>% add_markers(symbol = I(1), name = "hollow")
)
```

In Figure \@ref(fig:scatterplots), hollow circles are specified via `symbol = I(1)`. By default, the `symbol` argument (as well as the `color`/`size`/`linetype` arguments) assumes value(s) are "data", which need to be mapped to a visual palette (provided by `symbols`). Wrapping values with the `I()` function notifies `plot_ly()` that these values should be taken "AsIs". If you compare the result of `plot(1:25, 1:25, pch = 1:25)` to Figure \@ref(fig:pch), you'll see that `plot_ly()` can translate R's plotting characters (pch), but you can also use [plotly.js' symbol syntax](https://plot.ly/r/reference/#scatter-marker-symbol), if you desire.

```{r pch, fig.cap = "Specifying symbol in a scatterplot"}
subplot(
  plot_ly(x = 1:25, y = 1:25, symbol = I(1:25), name = "pch"),
  plot_ly(mpg, x = ~cty, y = ~hwy, symbol = ~cyl, symbols = 1:3, name = "cyl")
)
```

When mapping a numeric variable to `symbol`, it creates only one trace, so no legend is generated. If you do want one trace per symbol, make sure the variable you're mapping is a factor, as Figure \@ref(fig:symbol-factor) demonstrates. When plotting multiple traces, the default plotly.js color scale will apply, but you can set the color of every trace generated from this layer with `color = I("black")`, or similar.

```{r symbol-factor, fig.cap = "Mapping symbol to a factor"}
p <- plot_ly(mpg, x = ~cty, y = ~hwy, alpha = 0.3) 
subplot(
  add_markers(p, symbol = ~cyl, name = "A single trace"),
  add_markers(p, symbol = ~factor(cyl), color = I("black"))
)
```

The `color` argument adheres to similar rules as `symbol`:

* If numeric, `color` produces one trace, but [colorbar](https://plot.ly/r/reference/#scatter-marker-colorbar) is also generated to aide the decoding of colors back to data values. The `colorbar()` function can be used to customize the appearance of this automatically generated guide. The default colorscale is viridis, a perceptually-uniform colorscale (even when converted to black-and-white), and perceivable even to those with common forms of color blindness [@viridis].

* If discrete, `color` produces one trace per value, meaning a [legend](https://plot.ly/r/reference/#layout-legend) is generated. If an ordered factor, the default colorscale is viridis [@viridisLite]; otherwise, it is the "Set2" palette from the __RColorBrewer__ package [@RColorBrewer]

```{r, color-types, fig.cap = "Variations on a numeric color mapping."}
p <- plot_ly(mpg, x = ~cty, y = ~hwy, alpha = 0.5)
subplot(
  add_markers(p, color = ~cyl, showlegend = FALSE) %>% 
    colorbar(title = "Viridis", len = 1/2, y = 1),
  add_markers(p, color = ~factor(cyl))
) %>% layout(showlegend = TRUE)
```

There are a number of ways to alter the default colorscale via the `colors` argument. This argument excepts: (1) a color brewer palette name (see the row names of `RColorBrewer::brewer.pal.info` for valid names), (2) a vector of colors to interpolate, or (3) a color interpolation function like `colorRamp()` or `scales::colour_ramp()`. Although this grants a lot of flexibility, one should be concious of using a sequential colorscale for numeric variables (& ordered factors) as shown in \@ref(fig:color-numeric), and a qualitative colorscale for discrete variables as shown in \@ref(fig:color-discrete). (TODO: touch on lurking variables?)

```{r color-numeric, fig.cap = "Three variations on a numeric color mapping"}    
subplot(
  add_markers(p, color = ~cyl, colors = c("#132B43", "#56B1F7")) %>%
    colorbar(title = "ggplot2 default", len = 1/3, y = 1),
  add_markers(p, color = ~cyl, colors = viridisLite::inferno(10)) %>% 
    colorbar(title = "Inferno", len = 1/3, y = 2/3),
  add_markers(p, color = ~cyl, colors = colorRamp(c("red", "white", "blue"))) %>% 
    colorbar(title = "colorRamp", len = 1/3, y = 1/3)
)
```

```{r color-discrete, fig.cap = "Three variations on a discrete color mapping"}         
subplot(
  add_markers(p, color = ~factor(cyl), colors = "Pastel1"),
  add_markers(p, color = ~factor(cyl), colors = colorRamp(c("red", "blue"))),
  add_markers(p, color = ~factor(cyl), 
              colors = c(`4` = "red", `5` = "black", `6` = "blue", `8` = "green"))
) %>% layout(showlegend = FALSE)
```

For scatterplots, the `size` argument controls the area of markers (unless otherwise specified via [sizemode](https://plot.ly/r/reference/#scatter-marker-sizemode)), and _must_ be a numeric variable. The `sizes` argument controls the minimum and maximum size of circles, in pixels:

```{r}
subplot(
  add_markers(p, size = ~cyl, name = "default"),
  add_markers(p, size = ~cyl, sizes = c(1, 500), name = "custom")
)
```

#### 3D scatterplots

To make a 3D scatterplot, just add a `z` attribute:

```{r 3D-scatterplot, fig.cap = "A 3D scatterplot"}
plot_ly(mpg, x = ~cty, y = ~hwy, z = ~cyl) %>%
  add_markers(color = ~cyl)
```

#### Scatterplot matrices

I currently recommend creating scatterplot matrices via the `ggpairs()` function from the **GGally** package [@GGally] and using the `ggplotly()` function to convert it.

```{r, fig.asp = 1, fig.width = 8}
pm <- GGally::ggpairs(iris)
ggplotly(pm)
```

### Dotplots & error bars

A dotplot is similar to a scatterplot, except instead of two numeric axes, one is categorical. The usual goal of a dotplot is to compare value(s) on a numerical scale over numerous categories. In this context, dotplots are preferrable to pie charts since comparing position along a common scale is much easier than comparing angle or area [@graphical-perception]; [@crowdsourcing-graphical-perception]. Furthermore, dotplots can be preferrable to bar charts, especially when comparing values within a narrow range far away from 0 [@few-values]. Also, when presenting point estimates, and uncertainty associated with those estimates, bar charts tend to exaggerate the difference in point estimates, and lose focus on uncertainty [@messing].

A popular application for dotplots (with error bars) is the so-called "coefficient plot" for visualizing the point estimates of coefficients and their standard error. The `coefplot()` function in the **coefplot** package [@coefplot] and the `ggcoef()` function in the **GGally** both produce coefficient plots for many types of model objects in R using **ggplot2**, which we can translate to plotly via `ggplotly()`. Since these packages use points and segments to draw the coefficient plots, the hover information is not the best, and it'd be better to use [error objects](https://plot.ly/r/reference/#scatter-error_x). Figure \@ref(fig:coefplot) uses the `tidy()` function from the **broom** package [@broom] to obtain a data frame with one row per model coefficient, and produce a coefficient plot with error bars along the x-axis. 

```{r coefplot, fig.cap = "A coefficient plot"}
m <- lm(Sepal.Length ~ Sepal.Width * Petal.Length * Petal.Width, data = iris)
# arrange by estimate, then make term a factor to order categories in the plot
d <- broom::tidy(m) %>% 
  arrange(desc(estimate)) %>%
  mutate(term = factor(term, levels = term))
plot_ly(d, x = ~estimate, y = ~term) %>%
  add_markers(error_x = ~list(value = std.error)) %>%
  layout(margin = list(l = 200))
```

### Line plots

This section surveys useful applications of `add_lines()` and `add_paths()`. The only difference between these functions is that `add_lines()` connects x/y pairs from left to right, instead of the order in which the data appears. Both functions understand the `color`, `linetype`, and `alpha` attributes^[plotly.js currently [does not support data arrays for `scatter.line.width` or `scatter.line.color`](https://github.com/plotly/plotly.js/issues/147), meaning a single line trace can only have one width/color in 2D line plot, and consequently numeric `color`/`size` mappings won't work], as well as groupings defined by `group_by()`.

Figure \@ref(fig:houston) uses `group_by()` to plot one line per city in the `txhousing` dataset using a _single_ trace. Since there can only be one tooltip per trace, hovering over that plot does not reveal useful information. Although plotting many traces can be computationally expensive, it is necessary in order to display better information on hover. Since the `color` argument produces one trace per value (if the variable (`city`) is discrete), hovering on Figure \@ref(fig:many-traces) reveals the top ~10 cities at a given x value. Since 46 colors is too many to perceive in a single plot, Figure \@ref(fig:many-traces) also restricts the set of possible `colors` to black. 

```{r many-traces}
plot_ly(txhousing, x = ~date, y = ~median) %>%
  add_lines(color = ~city, colors = "black", alpha = 0.2)
```

Generally speaking, it's hard to perceive more than 8 different colors/linetypes/symbols in a given plot, so sometimes we have to filter data to use these effectively. Here we use the **dplyr** package to find the top 5 cities in terms of average monthly sales (`top5`), then effectively filter the original data to contain just these cities via `semi_join()`. Once we have the data is filtered, mapping city to `color` or `linetype` is trivial. The color palette can be altered via the `colors` argument, and follows the same rules as [scatterplots](#scatterplots). The linetype palette can be altered via the `linetypes` argument, and accepts R's [`lty` values](https://github.com/wch/r-source/blob/e5b21d0397c607883ff25cca379687b86933d730/src/library/graphics/man/par.Rd#L726-L743) or plotly.js [dash values](https://plot.ly/r/reference/#scatter-line-dash).

```{r}
library(dplyr)
top5 <- txhousing %>%
  group_by(city) %>%
  summarise(m = mean(sales, na.rm = TRUE)) %>%
  arrange(desc(m)) %>%
  top_n(5)

p <- semi_join(txhousing, top5) %>%
  plot_ly(x = ~date, y = ~median)

subplot(
  add_lines(p, color = ~city),
  add_lines(p, linetype = ~city),
  shareY = TRUE
)
```


#### Parallel Coordinates

One very useful, but often overlooked, visualization technique is the parallel coordinates plot. Parallel coordinates provide a way to compare values along a common (or non-aligned) positional scale(s) -- the most basic of all perceptual tasks -- in more than 3 dimensions [@graphical-perception]. Usually each line represents every measurement for a given row (or observation) in a data set. When measurements are on very different scales, some care must be taken, and variables must transformed to be put on a common scale. As Figure \@ref(fig:pcp-common) shows, even when variables are measured on a similar scale, it can still be a informative to transform variables in different ways.

```{r pcp-common, fig.width = 8, fig.cap = "Parallel coordinates plots of the Iris dataset. On the left is the raw measurements. In the middle, each variable is scaled to have mean of 0 and standard deviation of 1. On the right, each variable is scaled to have a minimum of 0 and a maximum of 1."}
iris$obs <- seq_len(nrow(iris))
iris_pcp <- function(transform = identity) {
  iris[] <- purrr::map_if(iris, is.numeric, transform)
  tidyr::gather(iris, variable, value, -Species, -obs) %>% 
    group_by(obs) %>% 
    plot_ly(x = ~variable, y = ~value, color = ~Species) %>% 
    add_lines(alpha = 0.3)
}
subplot(
  iris_pcp(), 
  iris_pcp(scale),
  iris_pcp(scales::rescale)
) %>% hide_legend()
```

It is also worth noting that the **GGally** offers a `ggparcoord()` function which creates parallel coordinate plots via **ggplot2**, which we can convert to plotly via `ggplotly()`. In [linked highlighting](#linked-highlighting), parallel coordinates are linked to lower dimensional (but sometimes higher resolution) graphics of related data to guide multi-variate data exploration.

#### 3D line plots

To make a 3D line plot, just add a `z` attribute:

```{r 3D-lines, fig.cap = "A 3D scatterplot"}
plot_ly(mpg, x = ~cty, y = ~hwy, z = ~cyl) %>%
  add_lines(color = ~displ)
```

### Segments

The `add_segments()` function essentially provides a way to connect two points ((`x`, `y`) to (`xend`, `yend`)) with a line. Segments form the building blocks for many useful chart types, including candlestick charts, a popular way to visualize stock prices. Figure \@ref(fig:candlestick) uses the **quantmod** package [@quantmod] to obtain stock price data for Microsoft and plots two segments for each day: one to encode the opening/closing values, and one to encode the daily high/low.

```{r candlestick, fig.cap = "A candelstick chart"}
library(quantmod)
msft <- getSymbols("MSFT", auto.assign = F)
dat <- as.data.frame(msft)
dat$date <- index(msft)
dat <- subset(dat, date >= "2016-01-01")

names(dat) <- sub("^MSFT\\.", "", names(dat))

plot_ly(dat, x = ~date, xend = ~date, color = ~Close > Open, 
        colors = c("red", "forestgreen"), hoverinfo = "none") %>%
  add_segments(y = ~Low, yend = ~High, size = I(1)) %>%
  add_segments(y = ~Open, yend = ~Close, size = I(3)) %>%
  layout(showlegend = FALSE, yaxis = list(title = "Price")) %>%
  rangeslider()
```

### Ribbons

Ribbons are useful for showing uncertainy bounds as a function of x. The `add_ribbons()` function creates ribbons and requires the arguments: `ymin` and `ymax`.

```{r}
m <- lm(mpg ~ wt, data = mtcars)
broom::augment(m) %>%
  plot_ly(x = ~wt, showlegend = FALSE) %>%
  add_markers(y = ~mpg, color = I("black")) %>%
  add_ribbons(ymin = ~.fitted - 1.96 * .se.fit, 
              ymax = ~.fitted + 1.96 * .se.fit, color = I("gray80")) %>%
  add_lines(y = ~.fitted, color = I("steelblue"))
```

### Polygons

The `add_polygons()` function is essentially equivalent to `add_paths()` with the [fill](https://plot.ly/r/reference/#scatter-fill) attribute set to "toself". Polygons for the basis for other, higher-level, geometries such as `add_ribbons()`, but can be useful in their own right. 

```{r map-canada, "A map of Canada using the default cartesian coordinate system."}
map_data("world", "canada") %>%
  group_by(group) %>%
  plot_ly(x = ~long, y = ~lat, alpha = 0.2) %>%
  add_polygons(hoverinfo = "none", color = I("black")) %>%
  add_markers(text = ~paste(name, "<br />", pop), hoverinfo = "text", 
              color = I("red"), data = maps::canada.cities) %>%
  layout(showlegend = FALSE)
```

## Maps

### Using scatter traces

As shown in [polygons](#polygons), it is possible to create maps using plotly's default (cartesian) coordinate system, but plotly.js also has support for plotting [scatter traces](#scatter-traces) on top of either a [custom geo layout](https://plot.ly/r/reference/#layout-geo) or a [mapbox layout](https://plot.ly/r/reference/#layout-mapbox). Figure \@ref(fig:maps) compares the three different layout options in a single subplot.

```{r maps, fig.width = 8, fig.cap = "A few maps"}
dat <- map_data("world", "canada") %>% group_by(group)

map1 <- plot_ly(dat, x = ~long, y = ~lat) %>% 
  add_paths(size = I(1)) %>%
  add_segments(x = -100, xend = -50, y = 50, 75)

map2 <- plot_mapbox(dat, x = ~long, y = ~lat) %>% 
  add_paths(size = I(2)) %>%
  add_segments(x = -100, xend = -50, y = 50, 75) %>%
  layout(mapbox = list(zoom = 0,
      center = list(lat = ~median(lat), lon = ~median(long))
   ))

# geo() is the only object type which supports different map projections
map3 <- plot_geo(dat, x = ~long, y = ~lat) %>% 
  add_markers(size = I(1)) %>%
  add_segments(x = -100, xend = -50, y = 50, 75) %>%
  layout(geo = list(projection = list(type = "mercator")))

subplot(map1, map2) %>%
  subplot(map3, nrows = 2) %>% 
  hide_legend()
```

Any of the `add_*()` functions found under [scatter traces](https://cpsievert.github.io/plotly_book/scatter-traces.html) should work as expected on plotly-geo (initialized via `plot_geo()`) or plotly-mapbox (initialized via `plot_mapbox()`) objects. You can think of `plot_geo()` and `plot_mapbox()` as special cases (or more opiniated versions) of `plot_ly()`. For one, they won't allow you to mix scatter and non-scatter traces in a single plot object, which you probably don't want to do anyway. In order to enable Figure \@ref(fig:maps), plotly.js _can't_ make this restriction, but since we have `subplot()` in R, we _can_ make this restriction without sacrificing flexibility.

### Choropleths

In addition to scatter traces, plotly-geo objects can also create a [choropleth](https://plot.ly/r/reference/#choropleth) trace/layer. Figure \@ref(fig:us-density) shows the population density of the U.S. via a choropleth, and also layers on markers for the state center locations, using the U.S. state data from the **datasets** package [@base]. By simply providing a [`z`](https://plot.ly/r/reference/#choropleth-z) attribute, plotly-geo objects will try to create a choropleth, but you'll also need to provide [`locations`](https://plot.ly/r/reference/#choropleth-locations) and a [`locationmode`](https://plot.ly/r/reference/#choropleth-locationmode).

```{r us-density, fig.cap = "A map of U.S. population density using the `state.x77` data from the **datasets** package."}
density <- state.x77[, "Population"] / state.x77[, "Area"]

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  lakecolor = toRGB('white')
)

plot_geo() %>%
  add_trace(
    z = ~density, text = state.name,
    locations = state.abb, locationmode = 'USA-states'
  ) %>%
  add_markers(
    x = state.center[["x"]], y = state.center[["y"]], 
    size = I(2), symbol = I(8), color = I("white"), hoverinfo = "none"
  ) %>%
  layout(geo = g)
```

## Bars & histograms

plotly.js has both a "bar" and "histogram" trace type. In statistical graphics, the term histogram is reserved for numeric variables, but the histogram trace type can be used to display categorical data.

```{r}
subplot(
  plot_ly(mpg, x = ~cyl),
  plot_ly(mpg, x = ~factor(cyl))
)
```

## Boxplots



## 2D Distributions


## 3D surfaces

* 

## Annotations

The `add_annotations()` function
