# Two approaches, one object

## A case study of housing sales in Texas {#txhousing-case-study}

The **plotly** package depends on **ggplot2** which bundles a data set on monthly housing sales in Texan cities acquired from the [TAMU real estate center](http://recenter.tamu.edu/). After the loading the package, the data is "lazily loaded" into your session, so you may reference it by name:

```{r}
library(plotly)
txhousing
```

In attempt to understand house price behavior over time, we could plot `date` on x, `median` on y, and group the lines connecting these x/y pairs by `city`. Using **ggplot2**, we can _initiate_ a ggplot object with the `ggplot()` function which accepts a data frame and a mapping from data variables to visual aesthetics. By just initiating the object, **ggplot2** won't know how to geometrically represent the mapping until we add a layer to the plot via one of `geom_*()` (or `stat_*()`) functions (in this case, we want `geom_line()`). In this case, it is also a good idea to specify alpha transparency so that 5 lines plotted on top of each other appear as solid black, to help avoid overplotting.

```{r}
p <- ggplot(txhousing, aes(date, median)) +
  geom_line(aes(group = city), alpha = 0.2)
```

### The `ggplotly()` function {#ggplotly}

Now that we have a valid **ggplot2** object, `p`, the **plotly** package provides the `ggplotly()` function which converts a ggplot object to a plotly object. By default, it supplies the entire aesthetic mapping to the tooltip, but the `tooltip` argument provides a way to restrict tooltip info to a subset of that mapping. Furthermore, in cases where the statistic of a layer is something other than the identity function (e.g., `geom_bin2d()` and `geom_hex()`), relevant calculated values are also supplied to the tooltip. This provides a nice mechanism for decoding visual aesthetics (e.g., color) used to represent a measure of interest (e.g, count/value). Figure \@ref(fig:ggsubplot) demonstrates tooltip functionality for a number of scenarios, and uses `subplot()` function from the **plotly** package (discussed in more detail in [Arranging multiple views](#arranging-multiple-views)) to concisely display numerous interactive versions of ggplot objects.

```{r ggsubplot, fig.width = 8, fig.cap = "Monthly median house price in the state of Texas. The top row displays the raw data (by city) and the bottom row shows 2D binning on the raw data. The binning is helpful for showing the overall trend, but hovering on the lines in the top row helps reveal more detailed information about each city.", screenshot.alt = "screenshots/ggsubplot"}
subplot(
  p, ggplotly(p, tooltip = "city"), 
  ggplot(txhousing, aes(date, median)) + geom_bin2d(),
  ggplot(txhousing, aes(date, median)) + geom_hex(),
  nrows = 2, shareX = TRUE, shareY = TRUE,
  titleY = FALSE, titleX = FALSE
)
```

The `ggplotly()` function translates most things that you can do in **ggplot2**, but not quite everything. To help demonstrate the coverage, I've built a [plotly version of the ggplot2 docs](http://ropensci.github.io/plotly/ggplot2). This version of the docs displays the `ggplotly()` version of each plot in a static form (to reduce page loading time), but you can click any plot to view its interactive version. The next section demonstrates how to create plotly.js visualizations via the R package, without **ggplot2**, via the `plot_ly()` function. We'll then leverage those concepts to [extend `ggplotly()`](#extending-ggplotly).

### The `plot_ly()` interface

#### The Layered Grammar of Graphics

The cognitive framework underlying the `plot_ly()` interface draws inspiration from the layered grammar of graphics [@ggplot2-paper], but in contrast to `ggplotly()`, it provides a more flexible and direct interface to [plotly.js](https://github.com/plotly/plotly.js). It is more direct in the sense that it doesn't call **ggplot2**'s sometimes expensive plot building routines, and it is more flexible in the sense that data frames are not required, which is useful for visualizing matrices, as shown in [Get Started](#get-started). Although data frames are not required, using them is highly recommended, especially when constructing a plot with multiple layers or groups. 

When a data frame is associated with a **plotly** object, it allows us to manipulate the data underlying that object in the same way we would directly manipulate the data. Currently, `plot_ly()` borrows semantics from and provides special plotly methods for generic functions in the **dplyr** and **tidyr** packages [@dplyr]; [@tidyr]. Most importantly, `plot_ly()` recognizes and preserves groupings created with **dplyr**'s `group_by()` function.

```{r}
library(dplyr)
tx <- group_by(txhousing, city)
# initiate a plotly object with date on x and median on y
p <- plot_ly(tx, x = ~date, y = ~median)
# plotly_data() returns data associated with a plotly object
plotly_data(p)
```

Defining groups in this fashion ensures `plot_ly()` will produce at least one graphical mark per group.^[In practice, it's easy to forget about "lingering" groups (e.g., `mtcars %>% group_by(vs, am) %>% summarise(s = sum(mpg))`), so in some cases, you may need to `ungroup()` your data before plotting it.] So far we've specified `x`/`y` attributes in the plotly object `p`, but we have not yet specified the geometric relation between these x/y pairs. Similar to `geom_line()` in **ggplot2**, the `add_lines()` function connects (a group of) x/y pairs with lines in the order of their `x` values, which is useful when plotting time series as shown in Figure \@ref(fig:houston).

```{r houston, fig.cap = "Monthly median house price in Houston in comparison to other Texan cities.", screenshot.alt = "screenshots/houston"}
# add a line highlighting houston
add_lines(
  # plots one line per city since p knows city is a grouping variable
  add_lines(p, alpha = 0.2, name = "Texan Cities", hoverinfo = "none"),
  name = "Houston", data = filter(txhousing, city == "Houston")
)
```

The **plotly** package has a collection of `add_*()` functions, all of which inherit attributes defined in `plot_ly()`. These functions also inherit the data associated with the plotly object provided as input, unless otherwise specified with the `data` argument. I prefer to think about `add_*()` functions like a layer in **ggplot2**, which is slightly different, but related to a plotly.js trace. In Figure \@ref(fig:houston), there is a 1-to-1 correspondence between layers and traces, but `add_*()` functions do generate numerous traces whenever mapping a discrete variable to a visual aesthetic (e.g., [color](scatterplots-discrete-color)). In this case, since each call to `add_lines()` generates a single trace, it makes sense to `name` the trace, so a sensible legend entry is created.

In the first layer of Figure \@ref(fig:houston), there is one line per city, but all these lines belong a single trace. We _could have_ produced one trace for each line, but this is way more computationally expensive because, among other things, each trace produces a legend entry and tries to display meaningful hover information. It is much more efficient to render this layer as a single trace with missing values to differentiate groups. In fact, this is exactly how the group aesthetic is translated in `ggplotly()`; otherwise, layers with many groups (e.g., `geom_map()`) would be slow to render.

